---
title: "Project 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data
```{r}
# call packages
library(tidyverse)
library(openintro)
library(factoextra)
library(psych)
library(cluster)
library(GGally)


# save cia_factbook dataset
data(cia_factbook)
# display first six rows of dataset
head(cia_factbook)

# remove rows with NA values
cia_clean <- cia_factbook %>%
  na.omit()

# make categorical variable for country type based on life expectancy at birth
cia_clean$country_type <- as.factor(ifelse(cia_clean$life_exp_at_birth <= 66, 'Least Developed',
                          ifelse(cia_clean$life_exp_at_birth > 74, 'More Developed',
                          ifelse((cia_clean$life_exp_at_birth > 66 && cia_clean$life_exp_at_birth <= 74), 'Less Developed'))))

cia_clean
```

## Title and Introduction
Write a narrative introduction describing the dataset you have chosen, the variables and observations it contains, how it was acquired (cite your sources), and why it is interesting to you. Expand on potential relationships, trends, you may expect, if any. Also comment on any steps needed to make the data tidy and clean. 

*Perhaps one the most popular international news stories over the past couple weeks has been regarding birth rates and population change in countries across the globe. In countries like Japan, population growth rate has seen a steady decline, leading some to worry the nation may soon age and disappear. In contrast, Sub-Saharan African nations are set to see a six-fold increase in population. As a group we found the trends in global population and development to be highly interesting, leading us to the CIA Factbook. This dataset is sourced from the openintro library in R. It contains columns for the country name, country area, population, population growth rate, birth rate, death rate, life expectancy, internet users, infant mortality rate, maternal mortality rate, and net migration rate for the year 2014. The original dataset has many rows with NA values so it requires cleaning by removing those rows. In addition, a categorical variable "country type" can be made based off of the existing variable for life expectancy. This dataset is interesting because we will be able to compare and contrast different trends for countries based on their statistics. We find this particular dataset to be interesting because it allows us to analyze the interconnected relationships between development as a nation and various statistics. In particular, we would like to look at each country's population and birth/death rates, as well as their life expectancy and mortality rates. We expect to see higher life expectancy and lower mortality rates in more developed countries, and the opposite for less and least developed countries. As for birth and death rates, we expect to see lower rates for more developed countries, and higher ones for less and least developed countries.*


## Exploratory Data Analysis
Explore relationships between your main variables by creating a correlation matrix with univariate/bivariate graphs and correlation coefficients. 
- Write a supporting paragraph describing which pairs of variables are the most/less correlated and any relationships/trends that are apparent. 

```{r}
# <<<<<<< HEAD
# create scatter plot for maternal mortality rate and infant mortality rate
# =======
# comment
# >>>>>>> fae302e41af64bebbbeafa22170a7b0cef308fb6
ggplot(cia_clean, aes(x = maternal_mortality_rate, y = infant_mortality_rate)) +
  geom_point()
```

```{r}
# find correlation between infant and maternal mortality rates
cor(cia_clean$maternal_mortality_rate,cia_clean$infant_mortality_rate, use = "pairwise.complete.obs")
```

```{r}
# create dataframe from cia_clean with only numeric variables
cia_num <- cia_clean %>%
  select_if(is.numeric) 

# build a correlation matrix between all numeric variables
cor(cia_num, use = "pairwise.complete.obs")
```

```{r}
# create a heatmap with geom_tile
cor(cia_num, use = "pairwise.complete.obs") %>%
  # save as a data frame
  as.data.frame %>%
  # convert row names to an explicit variable
  rownames_to_column %>%
  # pivot so that all correlations appear in the same column
  pivot_longer(-1, names_to = "other_var", values_to = "correlation") %>%
  ggplot(aes(rowname, other_var, fill = correlation)) +
  # create heatmap with geom_tile
  geom_tile() +
  # change the scale to make the middle appear neutral
  scale_fill_gradient2(low="red",mid="white",high="blue") +
  # overlay values
  geom_text(aes(label = round(correlation,2)), color = "black", size = 4) +
  # add title and axis labels
  labs(title = "Correlation Matrix for cia_factbook Numeric Variables", x = "variable 1", y = "variable 2") + 
  # rotate x-axis labels to make them more readable
  theme(axis.text.x=element_text(angle=45,hjust=1)) 
```

```{r}
# create correlation matrix with univariate and bivariate graphs
pairs.panels(cia_num, 
             method = "pearson", # correlation coefficient method
             hist.col = "blue", # color of histogram 
             smooth = FALSE, density = FALSE, ellipses = FALSE)
```
*When analyzing the CIA Factbook dataset, we first wanted to verify which columns had the greatest relationships. Using the correlation matrix and heat map, we determined the variables with the strongest relationships. We thought these findings would be useful to decide which variables to use for further analysis in the project. Looking at the heatmap and correlation coefficients between the variables, we found that life expectancy at birth and infant mortality rate have the strongest negative correlation with a correlation coefficient of -0.89. This makes sense because if infant mortality is high, it is likely that the life expectancy of these infants would be low. The next highest correlation found was between maternal mortality rate and infant mortality rate with a correlation coefficient of 0.86. This also makes sense as generally, the factors that drive maternal mortality also drive infant mortality. There was also a strong negative correlation between maternal mortality rate and life expectancy at birth. Again, this aligns with previous observations as women generally have children at younger ages. Another reason these two correlations make sense is healthcare. Nations with good healthcare can generally support older populations with less mortality during childbirth. There were also high correlations found between internet users and area and internet users and population. However, these variables were not chosen for further analysis because these correlations and the fact that a higher population/country area leads to a higher number of internet users do not indicate anything about how developed a country is, which is the main goal of the clustering and classification portions of this project.*

## Clustering
Perform k-means or PAM clustering on at least three of your variables (3 is the bare minimum: using more/all of them will make this much more interesting)!  
- Remember that you can incorporate categorical variables and perform clustering based on Gower dissimilarities:  convert character  variables  to  factors  in  R,  generate  Gower’s dissimilarity matrix on the data, and do PAM clustering on the dissimilarities. 
- Include all relevant steps such as the number of clusters based on silhouette width. 
- Visualize the clusters by showing all pairwise combinations of variables colored by cluster assignment (using ggpairs) AND/OR Visualize the clusters in a 2-dimension plot, using fviz_cluster for example (you could combine this section with the next section about dimensionality reduction).  
- Discuss and interpret the clusters in terms of the original variables and observations, report some statistics about the clusters, what observation is at the center, the goodness of fit, etc. 

```{r}
# STEP 1: choose number of clusters

# select variables, scale, and create new dataframe
cia_var <- cia_clean %>% 
  select(-country_type, -country, -area, -internet_users, -net_migration_rate, -population, -population_growth_rate) %>%
  scale

# find optimal number of clusters using within sum-of-squares (wss) method
# minimize WSS while keeping a small number of clusters
fviz_nbclust(cia_var, pam, method = "wss")

# find optimal number of clusters using silhouette method
# check silhouette width
fviz_nbclust(cia_var, pam, method = "silhouette")
```
*After finding our variables of interest, the next step in our project was to cluster our data. To do that, we searched for the best number of clusters. Using the WSS method (which measures the "compactness" of the clustering and minimizes it), it seems that the optimal number of clusters for the PAM method is around 2 or 3. We then implemented the silhouette method (which measures the quality of a clustering and determines how well each object lies within its cluster by considering both the WSS and the between-sum-of-squares), determining that the optimal number of clusters was indeed 2. Having found a number of clusters, we then moved on to clustering the data using the PAM method.*

# PAM Clustering
```{r}
# STEP 2: Use PAM method for clustering select variables (life expectancy, birth rate, death rate, maternal mortality rate, infant mortality rate)

# apply a clustering algorithm
pam_results <- cia_var %>%
  pam(k = 2)

# save cluster assignment as a column in dataset
cia_pam <- cia_clean %>%
  mutate(cluster = as.factor(pam_results$clustering))
```

```{r}
# STEP 3: visualize clusters
# make a plot of data colored by final cluster assignment (infant vs maternal mortality rates)
cia_pam %>% 
  ggplot(aes(maternal_mortality_rate, infant_mortality_rate, color = cluster)) +
  geom_point(aes(shape = country_type)) +
  labs(title="Clustering of Maternal and Infant Mortality Rates")

# make a plot of data colored by final cluster assignment (death rate vs birth rate)
cia_pam %>% 
  ggplot(aes(birth_rate, death_rate, color = cluster)) +
  geom_point(aes(shape = country_type)) +
  labs(title="Clustering of Birth Rate and Death Rate")
```
*After clustering the data into 2 clusters, we then created two plots, one displaying the relationship between maternal mortality and infant mortality and the other displaying birth rate vs death rate. In each plot, data points are shaped by country_type and colored by cluster. When viewing the scatter plots for infant vs maternal mortality rates, the clustering looks pretty accurate in that it clusters less and more developed countries into cluster 1 and least developed countries into cluster 2. This pattern with the clustering is also evident in the scatter plot for death rate vs birth rate. It was also interesting to see on both plots that the less developed nations mainly exist on the border between clusters 1 and 2. This was an exciting observation as it displays how the two clusters represent developed and underdeveloped nations.*

```{r}
# STEP 4: evaluate clustering by calculating accuracy from cluster and country type
# compare the cluster and species
table(cia_pam$cluster, cia_pam$country_type)

# calculate percentage of accuracy
(47+42+83/177)
```
*The accuracy percentage of this clustering using the pam method is around 89.47%, when comparing the clustering with the actual country type variable from the dataset.*

```{r}
# STEP 5.1: interpret clustering by visualizing clusters by showing pairwise combinations of variables

# show all pairwise combinations of variables colored by cluster assignment using ggpairs
ggpairs(cia_pam, columns = c(3,4,5,7,8), aes(color = cluster))
```
*(ADD NOTABLE OBSERVATIONS)*
```{r}
# STEP 5.2: interpret clustering by creating summary statistic for each variable

# find means of each variable for each cluster
cia_pam %>%
  group_by(cluster) %>%
  summarise_at(c("birth_rate", "death_rate", "infant_mortality_rate", "life_exp_at_birth", "maternal_mortality_rate"), mean, na.rm = T)
```

# K-Means Clustering
```{r}
# STEP 1: choose number of clusters

# find optimal number of clusters using within sum-of-squares (wss) method
# minimize WSS while keeping a small number of clusters
fviz_nbclust(cia_3, kmeans, method = "wss")

# find optimal number of clusters using silhouette method
# check silhouette width
fviz_nbclust(cia_3, kmeans, method = "silhouette")
```
*Using the WSS method, it seems that the optimal number of clusters for the kmeans method is around 2 or 3. Using the silhouette method, it is determined that the optimal number of clusters is 2.*

```{r}
# STEP 2: Use kmeans to cluster selected variables

# use kmeans function to find 2 clusters
kmeans_results <- kmeans(na.omit(cia_var),2)

# show available components
names(kmeans_results)

# visualize data by final cluster assignment
fviz_cluster(kmeans_results, data = cia_var)

# save cluster assignment as a column in dataset
cia_kmeans <- cia_clean %>%
  mutate(cluster = as.factor(kmeans_results$cluster))
```

```{r}
# STEP 3: visualize clusters

# visualize data by final cluster assignment (death rate vs birth rate)
cia_kmeans %>%
  ggplot(aes(maternal_mortality_rate, infant_mortality_rate, color = cluster)) +
  geom_point(aes(shape = country_type)) +
  labs(title="Clustering of Maternal and Infant Mortality Rates")

# visualize data by final cluster assignment (infant vs maternal mortality rates)
cia_kmeans %>%
  ggplot(aes(maternal_mortality_rate, infant_mortality_rate, color = cluster)) +
  geom_point(aes(shape = country_type)) +
  labs(title="Clustering of Birth Rate and Death Rate")
```
*When viewing the scatter plot for infant vs maternal mortality rates, the kmeans method mostly clusters less and more developed countries into cluster 2 and least developed countries into cluster 1. This pattern with the clustering is also evident when the scatter plot displays death rate vs birth rate. This is the opposite clustering (1 vs 2) as the pam method.*

```{r}
# STEP 4: evaluate clustering by calculating accuracy from cluster and country type

# compare the cluster and country type
table(cia_kmeans$cluster, cia_kmeans$country_type)

# calculate percentage of accuracy
(47+43+83/177)
```
*The accuracy percentage of this clustering using the pam method is around 90.47%, when comparing the clustering with the actual country type variable from the dataset. This accuracy is slightly higher than the one provided by the pam method.*

```{r}
# STEP 5.1: interpret clustering by visualizing clusters by showing pairwise combinations of variables

# show all pairwise combinations of variables colored by cluster assignment using ggpairs
ggpairs(cia_kmeans, columns = c(3,4,5,7,8), aes(color = cluster))
```

```{r}
# STEP 5.2: interpret clustering by creating summary statistic for each variable

# find means of each variable for each cluster
cia_kmeans %>%
  group_by(cluster) %>%
  summarise_at(c("birth_rate", "death_rate", "infant_mortality_rate", "life_exp_at_birth", "maternal_mortality_rate"), mean, na.rm = T)
```

## 4. Dimensionality Reduction
Perform PCA on at least three of your variables (3 is the bare minimum: using more/all of them will make this much more interesting)! 
- Use a prcomp to find the principal components. 
- Include all relevant steps such as the number of PCs based on variation explained. 
- Visualize the observations using the first 2 PCs (using one of the fviz_ functions). 
- Interpret the principal components, focusing on what it means to score low/high on each PC you retain, and report the total variation in your dataset explained by these PCs. 
 
```{r}


```


## Classification and Cross-Validation
Use  a  classifier,  (logistic  regression  or  k-Nearest Neighbor) to predict a binary variable (response) based on other variables in your dataset. 
- Train the model to the entire dataset and then use it to get predictions for all observations. Build a ROC curve and check how well the classifier is doing with the value of AUC. 
- Perform k-fold cross-validation with this same classifier. Get the average performance across your k folds. 
- Discuss the results in a paragraph. How well does your classifier predict new observations? Do you see signs of overfitting? 
 
```{r}

```
 
 
## Formatting. 
Create the report using R Markdown, with headers for each section; include comments to the R code; include references (datasets, context). The final report is less than 20 pages. If working in a group, acknowledge how each member contributed to the project. 



